Robust oscillations in multi-cyclic Markov state models of biochemical clocks

### 2019-12-20 reply-reviewers.tex

\begin{document}

\thispagestyle{empty}
\noindent Dear Referees,

\vspace{1cm}

Thank you for taking the time to carefully read our manuscript and for your thoughtful comments and criticisms. The main revision we have made is to add a section to the paper (``Markov state models of biochemical oscillators") that describes in more detail the models that we use and why we, and others in the literature~\cite{Barato2017}, believe them to be good representations of oscillator dynamics for our purposes. We have also clarified terms which we realized were poorly defined, such as ``robustness". Please find below a detailed response to each of your reviews. We hope that with these changes you will recommend the work for publication in JCP.

@@

\section*{Point-by-point reply}

The referee's comments are copied in italics, with our responses below.

\subsection*{Referee 1}

\textit{This is a paper of prime scientific quality that addresses the robustness of oscillations in cyclical reaction networks that are decorated by arbitrary side cycles. The work is solid and beyond scientific reproach (and hence I placed it as the ``top 5\%"), but there is an issue with the presentation: Do the authors really think JCP is the right venue for this article? It is way too specialized to be understood by even a small fraction of JCP readers. I have 3 decades of experience as a quantum chemical physics scientists and I had tremendous difficulty reading the paper due to the very specialized jargon (had to read certain passages times and time again, and to make things worse, many symbols in equations are not even defined in the text (some are defined only in figure captions) - to fully appreciate the paper one has to really read the previous paper from the authors' lab (and I did not have the time, unfortunately, to do so). A concrete application to a biochemical cycle would also be highly recommended.}

Thanks for your high estimation of the quality of our work. To make the work more accessible to the broader JCP readership, we have added a new section (``Markov state models of biochemical oscillators") in which we (a) give more detail about how the models we consider represent biochemical oscillators, (b) provide a specific example of an oscillator, and (c) clearly define the various terms and observables related to these models that we deal with in the paper, such as ``robustness", ``affinity", ``coherence", and so forth. We have also carefully gone through the manuscript and made sure that we have defined symbols and described our methods for generating the data in all of our figures. We hope that with these clarifications the manuscript is appropriate for publication in JCP.

@@

\subsection*{Referee 2}

\textit{In ``Robust oscillations in multi-cyclic models of biochemical clocks" the authors analyze the number and period of coherent oscillations in a continuous-time Markov chain without detailed balance. They build on a previous recent work where analytical expressions for these quantities where derived in the case of a unicycle chain very far from equilibrium. In the present manuscript they study the modifications induced by adding side cycles-on top of randomizing the transition rates-applying a coarse-graining procedure that leads them back to the unicycle case. The work is interesting, and well-written in places, and the theoretical/numerical results look solid. Nevertheless, I would appreciate the authors to address the following questions to help me pronouncing on the overall validity and relevance of the manuscript.}

\textit{i) On the basis of Eqs (4)-(6) - already derived in a preceding work and valid at large affinities-one clearly understands that ``the rates decouple from each other and only contribute additively to the timescales". The authors claim that, as a result, ``the farther an oscillator operates from equilibrium, the more robust it will be to fluctuations in the rates ". I do not see how this last statement follows from the previous one. It would do so, if they showed that the coupling between the heterogenous rates always increases in modulus $C \gamma$. Can this be seen from the derivation of Eqs (4)-(6)?}

Thanks for raising this point, which arises from our unclear definition of ``robustness". It appears that you reasonably assumed that by robustness we meant a smaller deviation of the observables from the uniform case, i.e., smaller $|C \gamma|$.  In fact, it is not obvious from Eqs (4)-(6), as far as we have been able to tell, that coupling between the heterogenous rates always increases $|C \gamma|$. However, we intended a different definition of robustness: a more robust network is one in which the period and coherence change less when some details of the network are changed, specifically the arrangement of the rates. We have clarified this in a section that we added to the manuscript titled ``Markov state models of biochemical oscillators" in which we also give more details about the model and terminology used in the paper. Since the expressions in Eqs.~4 - 6 contain no spatial correlations between rates, changing the arrangement of the rates will not change $|C \gamma|$ and the oscillator is ``robust" at high affinity. A scenario in which the values of the rates change but the distribution from which they are drawn stays the same (i.e., the rates fluctuate) can be roughly thought of as a rearrangement, and hence the oscillator is also more robust to fluctuations. We have edited and clarified our language around ``robustness" and ``fluctuations" so that our claims are clear and made the latter point more explicit in the manuscript.

@@

\textit{ii) The authors refer several time to the robustness of oscillations without defining it precisely. This leaves room to objections to their statement that higher affinity brings higher robustness. For example, we could require that the number of coherent oscillations $R$ decreases the least possible with respect to its value $R_0$ in the homogeneous (unicycle) case. If so, from Fig. 3 (lower left panel) we see that the lower affinity gives the highest $R/R_0$, i.e. it is actually the most robust condition. The same can be concluded looking at the exact values of $R$ in Fig. 4.
The authors should thus specify which notion of robustness they are considering to make their statement sound.}

As mentioned above, we have now explicitly defined robustness at the beginning of the paper. We hope this clarifies our statements.

@@

\textit{iii) The terminology employed in the conclusion is somewhat unclear. A net variation in the affinity is called ``long-lived, global rate fluctuations". However, the ``quenched disorder" in the rates could already by described as `long-lived and global'. The authors maybe mean that a variation in affinity represents extrinsic or environmental noise, while 'local' inhomogeneities in the rates effectively describe intrinsic noise?}

We have changed our language in the conclusion to clarify that we mean, on the one hand, randomness in the rates that preserves the average affinity, and on the other, changes in the affinity such that all of the rates either increase or decrease.

@@

\textit{iv) Clearly, a stochastic chemical reaction network is different from the Markov chain considered by the authors. In particular, since the dynamics is ruled by the Chemical Master Equation, the transition rates (with e.g. mass action kinetics) are state dependent, in general, and not constant.
Therefore, the present model should be seen as an abstract one to describe chemistry. In which sense is this an effective description? What kind of relation is there between the real chemical affinity (i.e. the difference in nonequilibrium free energy) and the present affinity?
All in all, should the effects observed for this Markov chain carry over to chemical reaction networks with mass action kinetics?}

As we now describe in the revised version of our paper, the Markov models that we consider represent the emergent behavior of a chemical reaction network that generates oscillations, and takes in to account mass action kinetics implicitly. In these models, each vertex represents a collective state of the system. The rates along each of the edges represent the rates of a reaction such as KaiA-KaiC binding. These rates depend on the states of the system represented by the connected vertices which encode the concentrations of all species, and therefore take into account mass action kinetics. In other words, two transitions at different locations in the cycle may be KaiA-KaiC binding events, but will have different rates since the binding rate depends on the concentration of free KaiA in the system which may be different in the different states~\cite{Paijmans2017}. The chemical affinity over a cycle is given by the net turnover of ATP that occurs over one cycle; individual steps in the cycle are made (more or less) irreversible by coupling to ATP hydrolysis.

To what extent this is an effective description then depends on whether reaction networks that produce oscillations generate Markov models with topology similar to what we consider in this work. This is an important question which has not to our knowledge thus far been properly investigated despite the use of such models to derive analytical results related to the thermodynamic properties of biochemical oscillators~\cite{Barato2017}. We are currently investigating this question using a detailed computational model of KaiC as an starting point~\cite{Paijmans2017}, and our preliminary results suggest that simulations of that model can indeed be projected on to cycles such as the ones we study here, and that the rates that emerge can be well-characterized as random variables with some spatial correlations.

\textit{Typos: \\
``...R and T change by only a small fraction as $\mu$ in turned on..." $\to$ ``...is turned on..." \\
``...high energy dissipation makes these observables insensitive tomany parameters" $\to$ ``...to many..."}

Thanks for noticing these typos. We have corrected them.

\vspace{1cm}
\noindent Regards,
\vspace{1cm}

@@

\noindent Suriyanarayanan Vaikuntanathan (Corresponding Author, Email: \href{mailto:svaikunt@uchicago.edu}{svaikunt@uchicago.edu})

\noindent Clara del Junco

%\bibliography{/Users/claradeljuncooffice/Dropbox/Papers/my-library.bib}

%merlin.mbs apsrev4-1.bst 2010-07-25 4.21a (PWD, AO, DPC) hacked
%Control: key (0)
%Control: author (8) initials jnrlst
%Control: editor formatted (1) identically to author
%Control: production of article title (-1) disabled
%Control: page (0) single
%Control: year (1) truncated
%Control: production of eprint (0) enabled
\begin{thebibliography}{2}%
\makeatletter
\providecommand \@ifxundefined [1]{%
 \@ifx{#1\undefined}
}%
\providecommand \@ifnum [1]{%
 \ifnum #1\expandafter \@firstoftwo
 \else \expandafter \@secondoftwo
 \fi
}%
\providecommand \@ifx [1]{%
 \ifx #1\expandafter \@firstoftwo
 \else \expandafter \@secondoftwo
 \fi
}%
\providecommand \natexlab [1]{#1}%
\providecommand \enquote  [1]{``#1''}%
\providecommand \bibnamefont  [1]{#1}%
\providecommand \bibfnamefont [1]{#1}%
\providecommand \citenamefont [1]{#1}%
\providecommand \href@noop [0]{\@secondoftwo}%
\providecommand \href [0]{\begingroup \@sanitize@url \@href}%
\providecommand \@href[1]{\@@startlink{#1}\@@href}%
\providecommand \@@href[1]{\endgroup#1\@@endlink}%
\providecommand \@sanitize@url [0]{\catcode `\\12\catcode `\$12\catcode
  `\&12\catcode `\#12\catcode `\^12\catcode `\_12\catcode `\%12\relax}%
\providecommand \@@startlink[1]{}%
\providecommand \@@endlink[0]{}%
\providecommand \url  [0]{\begingroup\@sanitize@url \@url }%
\providecommand \@url [1]{\endgroup\@href {#1}{\urlprefix }}%
\providecommand \urlprefix  [0]{URL }%
\providecommand \Eprint [0]{\href }%
\providecommand \doibase [0]{http://dx.doi.org/}%
\providecommand \selectlanguage [0]{\@gobble}%
\providecommand \bibinfo  [0]{\@secondoftwo}%
\providecommand \bibfield  [0]{\@secondoftwo}%
\providecommand \translation [1]{[#1]}%
\providecommand \BibitemOpen [0]{}%
\providecommand \bibitemStop [0]{}%
\providecommand \bibitemNoStop [0]{.\EOS\space}%
\providecommand \EOS [0]{\spacefactor3000\relax}%
\providecommand \BibitemShut  [1]{\csname bibitem#1\endcsname}%
\let\auto@bib@innerbib\@empty
%</preamble>
\bibitem [{\citenamefont {Barato}\ and\ \citenamefont
  {Seifert}(2017)}]{Barato2017}%
  \BibitemOpen
  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.~C.}\ \bibnamefont
  {Barato}}\ and\ \bibinfo {author} {\bibfnamefont {U.}~\bibnamefont
  {Seifert}},\ }\href {\doibase 10.1103/PhysRevE.95.062409} {\bibfield
  {journal} {\bibinfo  {journal} {Phys. Rev. E}\ }\textbf {\bibinfo {volume}
  {95}},\ \bibinfo {pages} {062409} (\bibinfo {year} {2017})}\BibitemShut
  {NoStop}%
\bibitem [{\citenamefont {Paijmans}\ \emph {et~al.}(2017)\citenamefont
  {Paijmans}, \citenamefont {Lubensky},\ and\ \citenamefont {ten
  Wolde}}]{Paijmans2017}%
  \BibitemOpen
  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
  {Paijmans}}, \bibinfo {author} {\bibfnamefont {D.~K.}\ \bibnamefont
  {Lubensky}}, \ and\ \bibinfo {author} {\bibfnamefont {P.~R.}\ \bibnamefont
  {ten Wolde}},\ }\href {\doibase 10.1016/J.BPJ.2017.05.048} {\bibfield
  {journal} {\bibinfo  {journal} {Biophys. J.}\ }\textbf {\bibinfo {volume}
  {113}},\ \bibinfo {pages} {157} (\bibinfo {year} {2017})}\BibitemShut
  {NoStop}%
\end{thebibliography}%

### 2019-12-20 main.tex
added letter to eds; this is the version resubmitted to JCP

\begin{abstract}
Organisms often use cyclic changes in the concentration of chemicals species to precisely biological functions. Underlying these biochemical clocks are series of chemical reactions and transport processes, which are inherently stochastic in the small and crowded environment of a cell. Understanding the physical basis for robust biochemical oscillations in the presence of fluctuations, as observed for instance in experiments of simple circadian oscillators, has emerged as an important problem. In a previous paper (arXiv:1808.04914), we explored this question using the non-equilibrium statistical mechanics of single-ring Markov state models of biochemical networks that support oscillations. Our finding was that they can exploit nonequilibrium driving to maintain a robust period and coherence in the presence of fluctuations in rates. Here, we extend our work to Markov state models consisting of a large cycle decorated with multiple small cycles. These additional cycles are intended to represent alternate pathways that the oscillator may take as it fluctuates about its average path. Combining a mapping to single-cycle networks based on first passage time distributions with our previously developed theory, we are able to make analytical predictions for the period and coherence of oscillations in these networks. One implication of our predictions is that for these networks, a high energy budget can make different network topologies and arrangements of rates degenerate as far as the period and coherence of oscillations is concerned. Excellent agreement between analytical and numerical results confirms that this is the case. Our results that biochemical oscillators can be more robust to fluctuations in the path of the oscillator when they have a high energy budget.
\end{abstract}

@@

Many organisms use internal biochemical clocks to synchronize their metabolism to day-night cycles, a tactic that confers fitness as it allows them to anticipate periodic environmental changes~\cite{Woelfle2004}. These clocks are implemented as a series of chemical reactions and transport processes, whose timing can be affected by intrinsic and extrinsic noise - yet typically these clocks have evolved to maintain consistent periods over different copies of the oscillator (e.g. in different cells), and over time. For example, the core clocks proteins of the circadian oscillator of {\it S. Elongatus} bacteria, generate oscillations in KaiC phosphorylation level with a 24-hour period over many days even in constant light or dark conditions~\cite{Nakajima2005, Tomita2005, Rust2007}.  From a theoretical standpoint, understanding the non-equilibrium statistical mechanical requirements for maintaining robust oscillations in these molecular clocks has thus emerged as an important question. In particular, the connection between energy dissipation and the precision of the stochastic period of oscillators has been noted by many theoretical studies~\cite{Barato2015, Cao2015, Barato2017, Fei2018, Wierenga2018, Nguyen2018, Marsland2019}. Some of this work proposes thermodynamic bounds which set a lower limit on the extent of stochastic fluctuations in these systems as a function of the energy dissipation budget~\cite{Barato2015, Barato2017, Wierenga2018}.  However, the structure of even simple models of biochemical oscillators constrains them to operate far from these bounds~\cite{Marsland2019}, which raises the question of what role energy dissipation plays in these cases where fluctuations are much larger than the minimum for the amount of energy that the oscillator is using.

Recently, we explored this question using single-cycle Markov models of biochemical oscillators such as that pictured in Fig.~\ref{fig:schematic}a. %operates when the rates are not symmetric, but rather chosen from a probability and subject to fluctuations, as is reasonably expected in any chemical reaction network operating in a noisy biological environment~\cite{DelJunco2018b}. 
By deriving an analytical expression for the period and coherence of oscillations that reveals their detailed dependence on all of the rates in the network, we showed that nonequilibrium driving allows the period of oscillators to become insensitive to many of the parameters of the models - specifically, the arrangement of the transition rates on the ring. Driving thus allows the period of a wide class of oscillators - even those operating far from the bound - to be robustly maintained in the presence of changes in these parameters. In this paper, we further explore this role of energy dissipation by extending our results to networks with multiple cycles. We show that the period of the oscillator is more robust at high driving to changes in the the topology of the network as well as the rates.

The paper is organized as follows: in Section~\ref{model} we introduce the class of multicyclic Markov state models considered in this paper and define nonequilibrium driving, the observables of interest (period and coherence of oscillations), and robustness in the context of these models. In Section \ref{theory} we briefly review our analytical theory from Ref.~\citenum{DelJunco2018b}. This theory depends on the single-cycle topology of the networks studied in Ref.~\citenum{DelJunco2018b}, so to apply it to multi-cycle networks such at the one illustrated in Fig.~\ref{fig:schematic}b, in Section~\ref{CG} we show how to coarse-grain small cycles on to single links, yielding an effective single-cycle network whose period and coherence are meant to approximate those of the full multi-cyclic network. In Section~\ref{timescales} we compare these observables calculated numerically in multi-cycle networks to the corresponding coarse-grained networks and analytical approximations, and show that at high affinity our analytical approximation, which takes as input only a small subset of the parameters required to specify the multi-cycle network, accurately reproduces the period and coherence in these networks. We demonstrate the ability of our theory to predict timescales when the rates and topology in the network are randomly generated. Finally, in Section~\ref{disc}, we discuss the implications of our results for biochemical oscillators and show one example of how the multi-cycle networks studied here can be used to achieve input compensation, which is the ability to maintain a constant period when the affinity changes.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig-1.pdf}
 \caption{A schematic of the networks studied in this article. (a) In previous work~\cite{DelJunco2018b}, we developed an analytical theory for the period of oscillations $T$ and the number of coherent oscillations $\R$, defined in Eq.~\ref{eq:r-t}, in oscillators that can be represented by a single cycle of states where the clockwise (CW) hopping rates ($k_i^+$) are much larger than the counterclockwise (CCW) hopping rates ($k_i^-$). This asymmetry is quantified by the affinity $\mathcal A$, defined in Eq.~\ref{eq:aff}. (b) In this work, we extend our results to networks where the main cycle is decorated with many small cycles. (c) We design these `decorations' so that the rates going into them are modulated by a small parameter $\mu$ which governs the probability that the system will enter the small cycle. (d) To apply our theory to these multi-cycle networks, we map the muti-cyclic network on to a unicyclic network by matching moments of the first passage time distribution for a random walk beginning CCW from the decoration (green circle) and ending CW from it (red circle) on to a line of states with two unknown hopping rate. This procedure gives a set of effective hopping rates $\eta_i^\pm$. When each of the decorations in the network is coarse-grained in this manner, we obtain a single-cycle network as in (a) whose period $T$ and coherence $\R$ should approximate those of the full, multicycle network.}
\label{fig:schematic}
\end{figure}

@@

In this paper we consider Markov models such as the one in Fig.~\ref{fig:schematic} as simple models that capture the cycling and stochasticity of biochemical oscillators~\cite{Barato2017}.  In this model, each vertex represents the collective state of a system. For instance, in the KaiABC oscillator, it could be a vector of the counts of each phosphorylation state of KaiC~\cite{Rust2007}. The rates along each of the edges represent the rates of elementary processes, like a phosphorylation event. We emphasize that this picture is thus not a representation of the underlying chemical reaction network which must contain, at a minimum, a negative feedback loop, and may also have other motifs~\cite{Novak2008}. Rather, it is an emergent picture that captures the oscillations that can arise from such a network, and the feedback as well as mass action kinetics are is encoded in the rates along each edge, which depend on the collective state of the system represented by the connected vertices. It is not expected that a real oscillator will always follow the same path through its state space on each cycle. The single-cycle model in Fig.~\ref{fig:schematic}a is a caricature that captures the average limit cycle of the oscillator. The multicycle model in Fig.~\ref{fig:schematic}b is a caricature intended to reflect of small fluctuations about this average.

The network is driven out of equilibrium by an affinity, $\mathcal A$, defined as 
\begin{equation}
\aff = \sum_{cycle} k_i^{+}/k_i^{-}.
\label{eq:aff}
\end{equation}
A finite affinity is necessary to have oscillations. The affinity is formally defined on closed cycles. In this work we refer to the affinity per site, $\aff/N = \langle k_i^{+}/k_i^{-} \rangle_{cycle}$, as a measure of the strength of driving. In biology, this nonequilibrium driving is typically provided by ATP hydrolysis. The quantities of interest are two time scales: the average period of oscillations $T$ and the number of coherent oscillations, $\R$, defined as
\begin{align}
T = 2\pi/ |\Im[\phi]| && \R = -|\Im[\phi]|/\Re[\phi] 
\label{eq:r-t}
\end{align}
where $\phi$ is the eigenvalue of the transition rate matrix of the network that yields the largest value of $\R$. \footnote{In refs.~\citenum{DelJunco2018b} and \citenum{Barato2017}, $\phi$ was defined as the eigenvalue with the least negative real part. However, in the multicyclic networks we address later in this paper, that definition can lead to selecting an eigenvalue which corresponds to cycling around a decoration rather than global oscillations.}. Loosely speaking, a higher value of $\R$ corresponds to smaller fluctuations in the period. Although $\R$ is one measure of the quality of timekeeping in the clock, it is not the definition of robustness that we use in this paper. In \citenum{Barato2017}, it was postulated that $\mathcal R$ is maximized in a uniform oscillator, that is, when all of the clockwise rates $k^+_i$ in the model in Fig.~\ref{fig:schematic}a are equal to one another and related to the counterclockwise rates $k^-_i$ by $k^+ = \exp(\aff/N) k^-$. However, since the rates along each edge of the network depend on the collective state of the system represented by the connected vertices, a network that sustains oscillations cannot be uniform~\cite{Marsland2019}. Moreover, the rates and, in the case of the multicyclic network depicted in Fig.\ref{fig:schematic}b, the locations and size of the small secondary cycles, can fluctuate over time and between copies of the oscillator. In this paper we therefore consider two properties of the oscillator: first, how predictable the period of oscillations $T$ and the coherence $\R$ are with limited knowledge about the specific details of the arrangement of rates and decorations in the network, and by extension, how robust these quantities are to fluctuations in these details.

@@

In Ref.~\citenum{DelJunco2018b}, we derived analytical expressions for $\phi$, and therefore the period and coherence of oscillations, in a network consisting of a single cycle of states, as depicted in Fig.~\ref{fig:schematic}a. In this section, we briefly review that result. Further details are available in Ref.~\citenum{DelJunco2018b}. An exact expression for $\phi$ in a cycle of $N$ states in the special symmetric case where $k_i^{+} = k^+$ and $k_i^{-} = k^-$ for all $i$ is given by:
\begin{equation}
 \phi^{(0)} = -(k^- + k^+) + k^-e^{-2\pi i/N} + k^+e^{2\pi i/N} \label{eq:phi0}.
\end{equation}

>>>>

\begin{abstract}
Organisms often use cyclic changes in the concentrations of chemicals species to precisely time biological functions. Underlying these biochemical clocks are chemical reactions and transport processes, which are inherently stochastic. Understanding the physical basis for robust biochemical oscillations in the presence of fluctuations has thus emerged as an important problem. In a previous paper (arXiv:1808.04914), we explored this question using the non-equilibrium statistical mechanics of single-ring Markov state models of biochemical networks that support oscillations. Our finding was that they can exploit non-equilibrium driving to robustly maintain the period and coherence of oscillations in the presence of randomness in the rates. Here, we extend our work to Markov state models consisting of a large cycle decorated with multiple small cycles. These additional cycles are intended to represent alternate pathways that the oscillator may take as it fluctuates about its average path. Combining a mapping to single-cycle networks based on first passage time distributions with our previously developed theory, we are able to make analytical predictions for the period and coherence of oscillations in these networks. One implication of our predictions is that a high energy budget can make different network topologies and arrangements of rates degenerate as far as the period and coherence of oscillations is concerned. Excellent agreement between analytical and numerical results confirms that this is the case. Our results suggest that biochemical oscillators can be more robust to fluctuations in the path of the oscillator when they have a high energy budget.
\end{abstract}

@@

Many organisms use internal biochemical clocks to synchronize their metabolism to day-night cycles, a tactic that confers fitness as it allows them to anticipate periodic environmental changes~\cite{Woelfle2004}. These clocks are implemented as a series of chemical reactions and transport processes, whose timing can be affected by intrinsic and extrinsic noise - yet they have evolved to maintain consistent periods over different copies of the oscillator (e.g. in different cells), and over time. For example, the circadian oscillator of {\it S. Elongatus} bacteria can be reconstituted {\it in vitro} from just 3 proteins, called KaiA, B, and C, which can sustain oscillations in the phosphorylation level of KaiC with a 24-hour period over many days even in constant light or dark conditions~\cite{Nakajima2005, Tomita2005, Rust2007}.  From a theoretical standpoint, understanding the non-equilibrium statistical mechanical requirements for maintaining robust oscillations in these molecular clocks has thus emerged as an important question. In particular, the positive connection between the amount of energy dissipated and the precision of the stochastic period of these oscillators has been noted by many theoretical studies~\cite{Barato2015, Cao2015, Barato2017, Fei2018, Wierenga2018, Nguyen2018, Marsland2019}. Some of this work proposes thermodynamic bounds which set a lower limit on the extent of stochastic fluctuations in these systems as a function of the energy dissipation budget~\cite{Barato2015, Barato2017, Wierenga2018}.  However, the structure of even simple models of biochemical oscillators constrains them to operate far from these bounds~\cite{Marsland2019}, which raises the question of what role energy dissipation plays in these cases where fluctuations are much larger than the minimum for the amount of energy that the oscillator is using.

Recently, we explored this question using single-cycle Markov models of biochemical oscillators such as that pictured in Fig.~\ref{fig:schematic}a~\cite{DelJunco2018b}. By deriving an analytical expression for the period and coherence of oscillations that reveals their detailed dependence on all of the rates in the network, we showed that non-equilibrium driving allows the period of oscillators to become insensitive to many of the parameters of the models - specifically, the arrangement of the transition rates on the ring. Driving thus allows the period of a wide class of oscillators - even those operating far from the bound - to be robustly maintained in the presence of changes in these parameters. In this paper, we further explore this role of energy dissipation by extending our results to networks with multiple cycles. We show that the period of the oscillator is more robust at high driving to changes in the the topology of the network as well as the rates.

The paper is organized as follows: in Section~\ref{model} we introduce the class of multi-cyclic Markov state models considered in this paper and define non-equilibrium driving, the observables of interest (period and coherence of oscillations), and robustness in the context of these models. In Section \ref{theory} we briefly review our analytical theory from Ref.~\citenum{DelJunco2018b}. This theory depends on the single-cycle topology of the networks studied in Ref.~\citenum{DelJunco2018b}, so to apply it to multi-cycle networks such as the one illustrated in Fig.~\ref{fig:schematic}b, in Section~\ref{CG} we show how to coarse-grain small cycles, which we call ``decorations", on to single links, yielding an effective single-cycle network whose period and coherence are meant to approximate those of the full multi-cyclic network. In Section~\ref{timescales} we compare these observables calculated numerically in multi-cycle networks to the corresponding coarse-grained networks and analytical approximations, and show that at high affinity our analytical approximation, which takes as input only a small subset of the parameters required to specify the multi-cycle network, accurately reproduces the period and coherence of these networks. We demonstrate the ability of our theory to predict timescales when the rates and topology in the network are randomly generated. Finally, in Section~\ref{disc}, we discuss the implications of our results for biochemical oscillators and show one example of how the multi-cycle networks studied here can be used to achieve input compensation, which is the ability to maintain a constant period when the affinity changes.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig-1.pdf}
 \caption{A schematic of the networks studied in this article. (a) In previous work~\cite{DelJunco2018b}, we developed an analytical theory for the period of oscillations $T$ and the number of coherent oscillations $\R$, defined in Eq.~\ref{eq:r-t}, in oscillators that can be represented by a single cycle of states where the clockwise (CW) hopping rates ($k_i^+$) are much larger than the counterclockwise (CCW) hopping rates ($k_i^-$). This asymmetry is quantified by the affinity $\mathcal A$, defined in Eq.~\ref{eq:aff}. (b) In this work, we extend our results to networks where the main cycle is decorated with many small cycles. (c) We design these `decorations' so that the rates going into them are modulated by a small parameter $\mu$ which governs the probability that the system will enter the decoration. (d) To apply our theory to these multi-cycle networks, we map the muti-cyclic network on to a unicyclic network by matching moments of the first passage time distribution for a random walk beginning CCW from the decoration (green circle) and ending CW from it (red circle) on to a line of states with two unknown hopping rates. This procedure gives a set of effective hopping rates $\eta_i^\pm$. When each of the decorations in the network is coarse-grained in this manner, we obtain a single-cycle network as in (a) whose period $T$ and coherence $\R$ approximate those of the full, multi-cycle network.}
\label{fig:schematic}
\end{figure}

@@

In this paper we consider Markov models, such as the ones in Fig.~\ref{fig:schematic}b, as simple models that capture the cycling and stochasticity of biochemical oscillators~\cite{Barato2017}.  In these models, each vertex represents a collective state of the system. For instance, in the KaiABC oscillator, it could be a vector of the counts of KaiC monomers in each phosphorylation state, and the concentrations of other species in solution~\cite{Rust2007}. The rates along each of the edges represent the rates of elementary processes, like a phosphorylation event. We emphasize that this picture is thus not a representation of the underlying chemical reaction network which must contain, at a minimum, a negative feedback loop, and may also have other motifs~\cite{Novak2008}. Rather, it is an emergent picture that captures the oscillations that can arise from such a network, and the feedback as well as mass action kinetics are encoded in the rates along each edge, which depend on the collective state of the system represented by the connected vertices. It is not expected that a real oscillator will always follow the same path through its state space on each cycle. The single-cycle model in Fig.~\ref{fig:schematic}a is a caricature that captures the average limit cycle of the oscillator. The multi-cycle model in Fig.~\ref{fig:schematic}b is a caricature intended to reflect small fluctuations about this average path.

The network is driven out of equilibrium by an affinity, $\mathcal A$, defined as 
\begin{equation}
\aff = \sum_{cycle} k_i^{+}/k_i^{-}.
\label{eq:aff}
\end{equation}
A finite affinity is necessary to have oscillations. The affinity is formally defined on closed cycles. In this work we refer to the affinity per site, $\aff/N = \langle k_i^{+}/k_i^{-} \rangle_{cycle}$, as a measure of the strength of driving. In biology, this non-equilibrium driving is typically provided by ATP hydrolysis and the affinity over the cycle will quantify the total net turnover of ATP per cycle. The quantities of interest are two time scales: the average period of oscillations $T$ and the number of coherent oscillations, $\R$, defined as
\begin{align}
T = 2\pi/ |\Im[\phi]| && \R = -|\Im[\phi]|/\Re[\phi] 
\label{eq:r-t}
\end{align}
where $\phi$ is the eigenvalue of the transition rate matrix of the network that yields the largest value of $\R$~\footnote{In refs.~\citenum{DelJunco2018b} and \citenum{Barato2017}, $\phi$ was defined as the eigenvalue with the least negative real part. However, in the multi-cyclic networks we address later in this paper, that definition can lead to selecting an eigenvalue which corresponds to cycling around a small decoration rather than global oscillations.}. Loosely speaking, a higher value of $\R$ corresponds to smaller fluctuations in the period. Although $\R$ is one measure of the quality of timekeeping in the clock, it is not the definition of robustness that we use in this paper. In \citenum{Barato2017}, it was postulated that $\mathcal R$ is maximized in a uniform oscillator, that is, when all of the clockwise (CW) rates $k^+_i$ in the model in Fig.~\ref{fig:schematic}a are equal to one another and related to the counterclockwise (CCW) rates $k^-_i$ by $k^+ = \exp(\aff/N) k^-$. However, since the rates along each edge of the network depend on the collective state of the system represented by the connected vertices, a network that sustains oscillations cannot be uniform~\cite{Marsland2019}. Moreover, the rates and, in the case of the multi-cyclic network depicted in Fig.\ref{fig:schematic}b, the locations and sizes of the small secondary cycles, can fluctuate over time and between copies of the oscillator. In this paper we therefore consider two properties of the oscillator: first, how predictable the period of oscillations $T$ and the coherence $\R$ are with limited knowledge about the specific details of the arrangement of rates and decorations in the network, and by extension, how robust these quantities are to fluctuations in these details.

@@

In Ref.~\citenum{DelJunco2018b}, we derived analytical expressions for $\phi$, and therefore the period and coherence of oscillations, in a network consisting of a single cycle of states, as depicted in Fig.~\ref{fig:schematic}a. In this section, we briefly review that result. Further details are available in Ref.~\citenum{DelJunco2018b}. An exact expression for $\phi$ in a cycle of $N$ states in the special uniform case where $k_i^{+} = k^+$ and $k_i^{-} = k^-$ for all $i$ is given by:
\begin{equation}
 \phi^{(0)} = -(k^- + k^+) + k^-e^{-2\pi i/N} + k^+e^{2\pi i/N} \label{eq:phi0}.
\end{equation}
We then considered networks where at least one of the CW rates is equal to $k^+$ and at least one of the CCW rates is equal to $k^-$. The remaining $m \leq N-1$ rates, denoted $h_j^\pm$, can be assigned arbitrary values ranging at least an order of magnitude above or below these ``uniform" rates. The main result of Ref.~\citenum{DelJunco2018b} was an expression for $\phi$ in this setup, in the limit of high affinity where terms of order $k^-/k^+ = \exp(-\aff/N)$ can be neglected compared to terms of order 1. The result is summarized in the following expressions: %which are the same as in Ref.~\citenum{DelJunco2018b} except for Eq.~\ref{eq:zeta} which has been modified to  relax some of the assumptions in Ref.~\citenum{DelJunco2018b} (further details in Appendix~\ref{app:theory}):
\begin{widetext}
\begin{align}
\phi &= \phi^{(0)} + C \gamma \label{eq:phinew} \\
 \phi^{(0)} & = -(k^- + k^+) + k^-\exp(-2 \pi i/N) + k^+\exp(2 \pi i/N) \\
\gamma &= \frac{1}{m-N}\left(\sum_{j=1}^m \log(\zeta_j(\gamma, k^\pm, h_j^\pm, N))\right) + \frac{1}{2(m-N)^2}\left(\sum_{j=1}^m \log(\zeta_j(\gamma, k^\pm, h_j^\pm, N))\right)^2 \label{eq:gamma} \\
%\zeta_j &=  \frac{{h_j^-} {k^+}+{h_j^+} {k^-}-{k^-} {k^+} + 2 \gamma  {h_j^+} {k^-} +\gamma ^2 \left({h_j^+} {k^-} + {k^-} {k^+} \right) + (\gamma +1) {k^+} e^{\frac{2 i \pi }{N}} (-{h_j^-}-{h_j^+}+{k^-}+{k^+})-\left((\gamma +1) {k^+} e^{\frac{2 i \pi }{N}}\right)^2}{(\gamma +1) {h_j^+} \left({k^-}-{k^+} e^{\frac{4 i \pi }{N}}\right)} \label{eq:zeta} \\
%C &=  c_1^2 k^- e^{-2\pi i/N} \label{eq:c} \\
%c_1^2 & = 1-(k^+/k^-)e^{4\pi i/N}. \label{eq:c1sq}
\end{align}
\end{widetext}
with expressions for $\zeta_j(\gamma, k^\pm, h_j^\pm, N))$ and $C$ given in Appendix~\ref{app:theory}. The essential feature of these equations is that Eq.~\ref{eq:gamma} depends independently on each rate $h_j^\pm$ and does not contain any information about the relative positions of the rates in the network. As a result, in the limit of moderately high affinity, spatial correlations vanish and the rates only contribute additively to the timescales, so we find that $\R$ and $T$ are insensitive to the arrangement of the rates in the network.  From a biological perspective, this means that the farther an oscillator operates from equilibrium, the more robust it will be to changes in the relative position of the rates. In a large enough network these rearrangements are akin to a scenario in which the values of the rates change but the distribution from which they are drawn stays the same (i.e., the rates fluctuate), and hence the oscillator is also more robust to fluctuations in the rates.

Since Eq.~\ref{eq:gamma} is a self-consistent equation for $\gamma$, the theoretical predictions in Section~\ref{timescales} are obtained numerically by searching for solutions to Eq.~\ref{eq:gamma} near to a linear approximation of Eq.~\ref{eq:gamma}.
